# Exercise 2: Employment in Europe

In this example we are going to look at a data set on employment in 26 European countries and perform PCA. 

The data gives for each of 26 European countries the percentage of the total workforce employed in nine different industries in 1979 (Hand et al, 1994).

Variable name   | Description
--------------- | ------------------------------------------
 Agriculture    | % employed in agriculture
 Mining         | % employed in mining
 Manufacture    | % employed in manufacturing
 Power          | % employed in power supply industries
 Construction   | % employed in construction
 Service        | % employed in service industries
 Finance        | % employed in finance
 Social         | % employed in social and personal services
 Transport      | % employed in transport & communications

```{r echo=TRUE}
employ<-read.table("eurojob.txt",header=T,row.names=1);head(employ,4)
```

**Task**

1. Perform exploratory data analysis (i.e. numerical summaries and some sensible plots) for these data. Comment on the results. 

`r hide("Hint")`
For numerical summaries, you could use `summary` or `skim`. For plots, you could use `pairs`.
`r unhide()`

2. Produce two important numerical summaries for deciding on how to run PCA and to tell how successful it is likely to be. Comment on these.

`r hide("Hint")`
Think about which statistics you will need in order to decide whether PCA might useful for this data, and whether to use covariance-based PCA or correlation-based PCA. 
`r unhide()`

3. Run PCA on the appropriate matrix and look at the output.

`r hide("Hint")`
Use the `princomp` command to implement PCA.
`r unhide()`

4. Assuming we are most concerned with preserving information, how many coefficients should we retain if we want to have 90\% of the original variability kept?

Answer: `r fitb(5)` components

5. Assuming we want to use Cattell's method, how many components would we retain?

`r hide("Hint")`
Use `plot` on your PCA model to produce a scree plot and look for the bend/elbow. 
`r unhide()`

Answer: `r fitb(c(2,4,6))` components

6. Assuming we want to use Kaiser's method, how many components would we retain?

Answer: `r fitb(3)` components

7. Assuming we have decided to retain 2 components, is there any useful interpretation to be had for these?

`r hide("Hint")`
Look at the loadings by using `$loadings`.
`r unhide()`

8. Produce a scatterplot of the data's scores for the first $2$ PCs and comment.

`r hide("Hint")``
PC scores are stored in `$scores` and a scatterplot can be produced by using `plot`. 
`r unhide()`

9. Say we have the following two new observations. 

```{r}
obs1<-c(5.1,0.5,32.3,0.8,8.1,16.7,4.3,21.2,6.3)
obs2<-c(4.2,0.7,25.4,0.7,9.3,15.0,5.8,31.0,6.9)
```

Calculate their scores on the second principal component. 

Answer: The second component score for `obs1` is `r fitb(c(-0.7608272,-0.760827,-0.76083,-0.7608,-0.761,-0.76))` and the second component score for `obs2` is `r fitb(c(-1.3729179,-1.372918,-1.37292,-1.3729,-1.373,-1.37))`.

`r hide("Hint")`
You will need to mean-centre and standardise both observations first and then multiply them by the second PC loadings.
`r unhide()`

